\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-03 tex/paper.tex
save
ref for VO

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

Variational optimization~\cite{2012arXiv1212.4507S} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
\end{equation}
where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
    %&= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
    &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
    &= \int \left[ f(\theta) \nabla_\psi \log q(\theta|\psi) \right]  q(\theta|\psi) d\theta \nonumber \\
    &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta) \nabla_\psi \log q(\theta|\psi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log
q(\theta|\psi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Density $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization. The proposal density correctly converges towards a distributon whose density concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

>>>>

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,staines2013optimization} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

Variational optimization~\cite{staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
\end{equation}
where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

Under mild restrictions outlined in  \citep{staines2013optimization}, the bound
$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
    %&= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
    &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
    &= \int \left[ f(\theta) \nabla_\psi \log q(\theta|\psi) \right]  q(\theta|\psi) d\theta \nonumber \\
    &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta) \nabla_\psi \log q(\theta|\psi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log
q(\theta|\psi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Density $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Density $q(\theta|\psi)$ after adversarial variational optimization ($\gamma=10$). The proposal density correctly converges towards a distributon whose density concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

\glnote{Discuss 1702.08235 and references therein on adversarial variational bayes/inference.}
